---
Title: "R for bioinformatics (and more!)"
menu: main
draft: false
date: false
---

Preview text for workshop Rmd

<!--more-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*R markdown version of content in R scripts*

## 2.3. `plotting.R`

### 2.3.1. Set Up

With the STRING data that we downloaded and previewed previously, we will create a network (aka. graph) representing protein interactions, and observe how certain proteins group together in "communities" with shared similarities.

```{r}
# load an R package that will be useful for our task
if (!require(igraph)) {
  install.packages("igraph")
  library(igraph)
}

# load data from STRING database, prepared in getData.R
load("./data/preparedData.RData")
```


### 2.3.2 Define functions
We will define a few functions that let us do certain operations on the data.

```{r}
getGraphSize <- function(G){
  # Find the difference in sizes between the largest 
  # and second largest communities in graph G
  # Parameters:
  #   G: an igraph
  #
  # Value: an integer
  
  # get cluster sizes
  sizes <- sizes(cluster_infomap(G))
  
  # order sizes as a vector
  sizesOrdered <- as.vector(sizes)
  
  # in some cases there is only one community
  if (length(sizesOrdered) == 1){
    # in this case, only return this community
    return(sizesOrdered[1])
  }
  
  # find difference between first and second largest community
  return(sizesOrdered[1] - sizesOrdered[2])
}

randGraph <- function(G){
  # Make a random graph of same degree as an input igraph G
  # Parameters:
  #   G: an igraph
  #
  # Value: an undirected igraph with same degree as G
  
  d <- degree(G)
  return(sample_degseq(d, method = "vl"))
}
```



### 2.3.3. Plot network

This sections is modified from Boris Steipe's [FND-MAT-Graphs_and_networks.R](https://github.com/hyginn/ABC-units) Version 1.0 

Create an igraph object from our STRING network, by plotting the network 
according to community membership of each protein (represented as nodes)

An edge in our network is defined between a pair of proteins. For all the pairs 
in our dataset, we want to create an edge. We select these as columns 1 and 2 of 
our prepared dataset. 


```{r}
stringNetwork <- graph_from_edgelist(as.matrix(preparedData[c(1,2)]), directed = F)

# With functions from iGraph package, prepare network for plotting
communities <- cluster_infomap(stringNetwork)
networkLayout <- layout_with_graphopt(stringNetwork, charge = 0.002, mass=100, spring.constant = 0.5)

# Plot!
plot( stringNetwork,
      layout = networkLayout,
      rescale = F,
      xlim = c(min(networkLayout[,1]) * 0.99, max(networkLayout[,1]) * 1.01),
      ylim = c(min(networkLayout[,2]) * 0.99, max(networkLayout[,2]) * 1.01),
      vertex.color=rainbow(max(membership(communities)+1))[membership(communities)+1],
      vertex.size = 700 + (90 * degree(stringNetwork)),
      vertex.label = NA,
      main = "Network from STRING data"
    )
```

With these settings, size of a node increases with degree, colour is dictated by 
community membership. We can see that there are several fairly well connected nodes 
with high degree of connectivity to other nodes. These nodes tend to be members of 
larger communities - this makes sense biologically, we could expect a protein in a 
large group of interacting proteins has many interactions. There are two distinctly 
large and well connected communities.

